\bf A Note on the Instability and Degeneracy of Deep Learning Models

### 2017-11-02 writing/paper-note/docs/note.tex
Revert "nordman's changes"
This reverts commit 2972c56368bdf5d74b94c5b34c0788d109a74bad.

\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{longtable,booktabs}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={A note on the instability and degeneracy of deep learning models},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{5}

@@

\title{\bf A note on the instability and degeneracy of deep learning models}
\author{Andee Kaplan \\ Department of Statistics, Iowa State University \\  and \\ Daniel Nordman \\ Department of Statistics, Iowa State University \\  and \\ Stephen Vardeman \\ Department of Statistics and Department of Industrial and\\
Manufacturing Systems Engineering, Iowa State University \\ }

@@

We consider the behavior, and the potential impropriety, sequences of
discrete probability models built to incorporate observations of
increasing sample size \(N\). Interest is in identifying instability in
such models, which is roughly characterized by probabilities with
extreme sensitivity to small changes in data configuration. The concept
of instability was introduced in the field of statistical physics (i.e.,
point processes) by Ruelle
(\protect\hyperlink{ref-ruelle1999statistical}{1999}) and then further
extended by Schweinberger Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011}) for a
family of exponential models. At issue, models exhibiting instability
are typically undesirable as these tend to provide poor representations
of data or data-generation. As an example, such models can include
near-degenerate distributions that assign all essential probability mass
to only a subset of an overall sample space. The latter issue in
connection to degeneracy has been recognized as concern in that dominant
model outcomes may not resemble observed data (cf. Handcock
\protect\hyperlink{ref-handcock2003assessing}{2003}). As a compounding
issue, model instability often has direct negative impacts for
statistical inference and computations based on likelihood functions.
Namely, volatilities in probability structure can potentially hamper the
numerical evaluations required for maximum likelihood estimation as well
as other model-based simulations via Markov Chain Monte Carlo (MCMC).
These reasons motivate our general study of instability for a broad
class of probability models, described next.

@@

Section \ref{examples} provides some examples of FOES models encountered
in graph/network analysis and machine learning (i.e., deep learning
models). These are used as references for later illustrations. Section
\ref{instability-results} then establishes several formal results for
FOES models with regard to instability. Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011}) originally
developed instability results framed for, or specific to, a certain
class of discrete exponential models. For similar exponential models
with random networks, Handcock
(\protect\hyperlink{ref-handcock2003assessing}{2003}) studied model
degeneracy, where a probability model places near complete mass on modes
and may thereby narrow the feasible model outcomes. As findings here and
from Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011}) suggest,
model instability and degeneracy may also be related by viewing
degeneracy as an extreme, or limiting form, of instability. Our main
results establish a broad characterization of model instability,
appropriate across the whole FOES model class, that incorporates results
of Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011}) as a
special case. We prescribe a general and simple condition for
identifying instability in a FOES model sequence, which quantifies
whether certain maximal probabilities in a FOES model are too extreme
relative to the sample size \(N\). When fulfilled, the probability
structure of a FOES model is shown to exhibit extreme sensitivity, with
probability assignments possessing extreme peaks and troughs across
nearly identical outcomes. As the measure of model instability
increases, probabilities from an unstable FOES model additionally
increase in volatility and provably slide into degeneracy. Section
\ref{implications} then emphasizes the implications of such model
instability, showing that such impropriety can be expected to
numerically hinder maximum likelihood estimation and MCMC-based
simulations. As one potential remedy, suggestions are given or
constraining model parameterizations and avoid the most problematic
regions of the parameter space. Proofs of the main results appear in
Appendix \ref{appendix-instab}.

@@

\subsection{Discrete exponential family
models}\label{discrete-exponential-family-models}

@@

\subsection{Restricted Boltzmann
machines}\label{restricted-boltzmann-machines}

@@

\(N_\mathcal{H}\) of hidden variables \(\boldsymbol H\) in the joint RMB

@@

\subsection{Deep learning}\label{deep-learning}

@@

\textbf{Deep belief network (DBN).} A DBN resembles a DBM in that there
are multiple layers of latent random variables stacked in a deep
architecture with no conditional dependence between layers. The
difference between the DBM and DBN models is that all but the last
stacked layer in a DBN are Bayesian networks (see Pearl
\protect\hyperlink{ref-pearl985bayesian}{1985}), rather than RBMs. Thus
for visibles \(X_1, \dots, X_N\) with support \(\mathcal{X}^N\), a DBN
is also a FOES model if the number \(q(N)\) of components in the
parameter vector is dependent on the dimension of the visibles.
Commonly, as in logistic belief nets (Neal
\protect\hyperlink{ref-neal1992connectionist}{1992}), a ``weight''
parameter is placed on each interaction between visibles,
\(X_1, \dots, X_N\), and the first layer of latent variables,
\(H^{(1)}_1, \dots, H^{(1)}_{N_{(H,1)}}\), in the definition of a FOES
model.

\section{Main results on model instability}\label{instability-results}

\subsection{A criterion for instability}\label{criterion}

To define a measure of instability in FOES models, it is useful to
consider the behavior of data models \(P_{\theta_N}\), again supported
on a set \(\mathcal{X}^N\) of outcomes for
\(\boldsymbol X\equiv \boldsymbol X_N =(X_1,\ldots,X_N)\), in connection
to the sample size \(N\). A relevant quantity to this end is a log-ratio
of extremal probabilities (LREP), defined as
\begin{align}
\label{eq:elpr}
 \REP (\boldsymbol \theta_N)  =  \log \left[\frac{\max\limits_{  \boldsymbol x\in \mathcal{X}^N}P_{\boldsymbol \theta_N}( \boldsymbol x)}{\min\limits_{ \boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}( \boldsymbol x)}\right],
\end{align}
based on maximum and minimal model probabilities. In what follows, the
main idea is that instability, and other negative model features, can be
associated with a FOES model formulation for \(N\) random variables
where the log-ratio \eqref{eq:elpr} is overly large relative to the sample
size \(N\). That is, a sequence of FOES probability models
\(P_{\boldsymbol \theta_N}\) results in specifying the distribution of
observations \(\boldsymbol X=(X_1,\ldots,X_N)\) for each sample size
\(N \geq 1\) and instability will generally occur among these models
whenever the corresponding LREP \eqref{eq:elpr} grows faster than \(N\).
This leads to the following definition.

@@

This definition of S-unstable is a generalization or reinterpretation of
``unstable'' used in Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011}) by allowing
possibility non-exponential family models (e.g., RBM and DBM models in
Sections \ref{restricted-boltzmann-machines} - \ref{deep-learning} as
well as a potentially increasing number \(d(N)\) of parameters through
the parameter sequence \(\boldsymbol \theta_N\in \mathbb{R}^{d(N)}\).
While this definition differs in form and scope from the original, it
does match that in Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011}) for the
special case of exponential models (cf.~Section
\ref{discrete-exponential-family-models} considered there. Section
\ref{illustrations} provides several examples of unstable models as well
as causes for model instability, where the latter may often be traced to
issues in model form (i.e., data functions) and/or parameterization. We
next describe several potentially undesirable features associated with
unstable models.

@@

\subsection{Characterizations and consequences of
instability}\label{characterizations-and-consequences-of-instability}

As a basic characteristic, S-unstable FOES model sequences have
extremely sensitive probability structures. One aspect is that small
changes in data configuration can lead to very large changes in
probability. Consider, for example, the quantity given by \[
\DN(\boldsymbol \theta_N) \equiv \max \left\{\log \frac{P_{\boldsymbol \theta_N}(\boldsymbol x)}{P_{\boldsymbol \theta_N}(\boldsymbol x^*)} : \boldsymbol x \text{ }\& \text{ } \boldsymbol x^* \in \mathcal{X}^N \text{ differ in exactly one component}\right\},
\] which represents the biggest log-probability ratio for a one
component change in data outcomes in a FOES model with parameter
\(\boldsymbol \theta_N\). We then have the following result prescribing
the behavior of \(\DN(\boldsymbol \theta_N)\) for unstable models.

@@

Theorem \ref{thm:instab-elpr}(i) is a non-asymptotic result, which
connects to the definition of instability in a FOES model through a
log-ratio of extreme model probabilities \eqref{eq:elpr} being too large
relative to the associated sample size \(N\). If so, Theorem
\ref{thm:instab-elpr}(i) guarantees the FOES model must also exhibit
correspondingly large changes in probability for very small differences
among some data configurations, a property that intuitively captures a
notion of instability. Furthermore, and perhaps more seriously under
Theorem \ref{thm:instab-elpr}(ii), S-unstable models can never have
universally bounded changes in probability among single component
variations in data configurations. While not all one component changes
in data may produce massive changes in probability, unstable models must
have some such data outcomes with this property. As a consequence,
unstable probability structure may exhibit extreme sensitivity through
large peaks and troughs over the sample space.

Additionally, S-unstable FOES model sequences are also connected to
degenerate models, where \emph{degeneracy} involves assigning
essentially all probability to modes within the sample space, which
could potentially represent a small subset among the totality of
outcomes. For perspective, note that differing sizes of the scaled
log-ratio \(\REP(\boldsymbol \theta_N)/N\) from \eqref{eq:elpr} induce a
spectrum of levels of instability/stability and Theorem
\ref{thm:instab-elpr} indicates increasing sensitivity of model
probabilities as \eqref{eq:elpr} increases. Furthermore, as the
instability measure grows and the log-ratio
\(\REP(\boldsymbol \theta_N)/N\) diverges, as in the definition
\eqref{eq:Sun} of S-unstable models, then a FOES model sequence will
become degenerate. Theorem \ref{thm:degenFOES} provides a formal
statement of such degeneracy due to instability. For a given
\(0 < \epsilon < 1\), define a \(\epsilon\)-modal set of outcomes as
\begin{equation}
\label{eq:mode}
\mathcal{M}_{\epsilon, \boldsymbol \theta_N} \equiv \left\{\boldsymbol x \in \mathcal{X}^N: \log P_{\boldsymbol \theta_N}(\boldsymbol x) > (1-\epsilon)\max\limits_{\boldsymbol y \in \mathcal{X}^N} \log  P_{\boldsymbol \theta_N}(\boldsymbol y) + \epsilon\min\limits_{\boldsymbol y \in \mathcal{X}^N} \log P_{\boldsymbol \theta_N}(\boldsymbol y) \right\}.
\end{equation}
\BeginKnitrBlock{theorem}
\protect\hypertarget{thm:degenFOES}{}{\label{thm:degenFOES}}For any
arbitrarily small \(0 < \epsilon < 1\), an S-unstable FOES model
sequence \(P_{\boldsymbol \theta_N}\), \(N \geq 1\), for
\(\boldsymbol X_N=(X_1, \dots, X_N)\) satisfies \[
P_{\boldsymbol \theta_N}\left( \boldsymbol X_N\in \mathcal{M}_{\epsilon, \boldsymbol \theta_N}\right) \rightarrow 1 \text{ as } N \rightarrow \infty.
\]
\EndKnitrBlock{theorem}

@@

Connected to degeneracy, unstable models may also exhibit additional
kinds of extreme and undesirable sensitivity in probabilities if model
parameters \(\boldsymbol \theta_N\) can further be ``dialed'' between
positive and negative values. That is, some FOES models naturally
involve parameter spaces covering a positive-negative spectrum of
parameter possibilities, where the signs of parameters provide a
standard device for increasing or decreasing probabilities of outcomes
in the model formulation. In fact, for many models, the switch of a
parameter sign serves to produce reciprocal probabilities, as outlined
in the following model assumption about parameter sign reversal (PSR).

@@

\BeginKnitrBlock{remark}[need to briefly finish]
\iffalse <span class="remark"><em>Remark (need to briefly finish). \fi{}
extensions of sensitivity\ldots{} basically, model condition PSR and
Corollary \ref{cor:sign} can be extended to cases where parameter
components
\(\boldsymbol \theta_N =(\boldsymbol \theta_{1,N}, \boldsymbol \theta_{N,2})\)
are not all changed in sign (\(-\boldsymbol \theta_N\)) but, more
generally, are instead altered to another parameter configuration
\(\boldsymbol \theta_{N}^A = ( \boldsymbol \theta_{1,N}^A, \boldsymbol \theta_{2,N}^A )\)
involving a switch in sign only among some model parameters
\(\boldsymbol \theta_{N,2}^A=- \boldsymbol \theta_{N,2}\) with remaining
parameters \(\boldsymbol \theta_{1,N}^A\) being arbitrarily chosen. If a
change sign occurs among parameters (\(\pm \boldsymbol \theta_{N,2}\))
which dominate the probability structure of the model, then an unstable
model under \(\boldsymbol \theta_N\) will then imply that many more
unstable models exist over a broad spectrum of possibilities for
variations \(\boldsymbol \theta_{N}^A\) of \(\boldsymbol \theta_N\).
\EndKnitrBlock{remark}

@@

Model instability can depend intricately on how functions of parameters
and data \(\boldsymbol X_N=(X_1,\ldots,X_n)\) are combined in the
formulation of the model probabilities, though some general causes may
be identified. As one issue, a broad parameter space (or wide
interpretation of this space) may admit some parameters as technically
valid that have an undue and often undesirable impact on the model
structure for a prescribed data size \(N\). In this case, both the size
and dimension of model parameters can be problematic and induce
instability. In combination to this last point, further causes of
instability may also be traced to the magnitude of statistics in the
model. Potentially massive, and thereby unstable, statistics were the
primary focus of instability studies of Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011}) for certain
discrete exponential models having parameters/statistics of fixed
dimension. However, as shown in the following, bounded statistics may
still lead to instability if the parameter dimension is high. We next
provides some examples to illustrate instability in FOES models, which
also suggest some potential strategies for preventing unstable models.

\subsection{Equi-probability models}\label{equi-probability-models}

@@

\subsection{One-parameter exponential models}\label{one-param-exp}

@@

\subsection{Fixed-dimensional linear exponential
models}\label{fixed-dim-exp}

@@

\(P_{\boldsymbol \theta_N}\) denote a RBM-based model \eqref{eq:RBM2} for
\(N\geq 1\) visible variables \(\boldsymbol X \equiv \boldsymbol X_N\)
derived from \(\tilde{P}_{\boldsymbol \theta_N}\) as the joint RBM
distribution \eqref{eq:RBM1} of \((\boldsymbol X, \boldsymbol H)\)
involving some number \(N_{\mathcal{H}} \equiv N_{N,\mathcal{H}}\geq 0\)
of hidden variables \(\boldsymbol H \equiv \boldsymbol H_N\) and
\(\boldsymbol \theta_N \equiv (\boldsymbol \theta_N^{\mathcal{V}},\boldsymbol \theta_N^{\mathcal{H}}, \boldsymbol \theta_N^{\mathcal{VH}}) \in\mathbb{R}^{N}\times \mathbb{R}^{N_{\mathcal{H}}} \times \mathbb{R}^{N N_{\mathcal{H}}}\).

@@

\item the instability measure $\REP_{(\boldsymbol X)}(\boldsymbol \theta_N)$ for the marginal model $P_{\boldsymbol \theta_N}$ of  $\boldsymbol X$ satisfies

@@

    \left| \REP_{(\boldsymbol X)}(\boldsymbol \theta_N)  - \elt\right| \leq    N_{\mathcal{H}}  \log 2

@@

\subsection{Implications for
simulation}\label{implications-for-simulation}

>>>>

\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{longtable,booktabs}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={A Note on the Instability and Degeneracy of Deep Learning Models},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{5}

@@

\title{\bf A Note on the Instability and Degeneracy of Deep Learning Models}
\author{Andee Kaplan \\ Department of Statistics, Iowa State University \\  and \\ Daniel Nordman \\ Department of Statistics, Iowa State University \\  and \\ Stephen Vardeman \\ Department of Statistics and Department of Industrial and\\
Manufacturing Systems Engineering, Iowa State University \\ }

@@

We consider the behavior, and the potential impropriety, of sequences of
discrete probability models built to incorporate observations of
increasing sample size \(N\). Interest is in identifying instability in
such models, which is roughly characterized by probabilities with
extreme sensitivity to small changes in data configuration. The concept
of instability was introduced in the field of statistical physics (i.e.,
point processes) by Ruelle
(\protect\hyperlink{ref-ruelle1999statistical}{1999}) and then further
extended by Schweinberger Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011}) for a
family of exponential models. At issue, models exhibiting instability
are typically undesirable as these tend to provide poor representations
of data or data-generation. As an example, such models can include
near-degenerate distributions that assign all essential probability mass
to only a subset of an overall sample space. The latter issue in
connection to degeneracy has been recognized as a concern in that
dominant model outcomes may not resemble observed data (cf. Handcock
\protect\hyperlink{ref-handcock2003assessing}{2003}). As a compounding
issue, model instability often has direct negative impacts for
statistical inference and computations based on likelihood functions.
Namely, volatilities in probability structure can potentially hamper the
numerical evaluations required for maximum likelihood estimation as well
as other model-based simulations via Markov Chain Monte Carlo (MCMC).
These reasons motivate our general study of instability for a broad
class of probability models, described next.

@@

Section \ref{examples} provides some examples of FOES models encountered
in graph/network analysis and machine learning (i.e., deep learning
models). These are used as references for later illustrations. Section
\ref{instability-results} then establishes several formal results for
FOES models with regard to instability. Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011}) originally
developed instability results specific to a certain class of discrete
exponential models. For similar exponential models with random networks,
Handcock (\protect\hyperlink{ref-handcock2003assessing}{2003}) studied
model degeneracy, where a probability model places near complete mass on
modes and may thereby narrow the feasible model outcomes. As findings
here and from Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011}) suggest,
model instability and degeneracy may also be related by viewing
degeneracy as an extreme, or limiting form, of instability. Our main
results establish a broad characterization of model instability,
appropriate across the whole FOES model class, that incorporates results
of Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011}) as a
special case. We prescribe a general and simple condition for
identifying instability in a FOES model sequence, which quantifies
whether certain maximal probabilities in a FOES model are too extreme
relative to the sample size \(N\). When fulfilled, the probability
structure of a FOES model is shown to exhibit extreme sensitivity, with
probability assignments possessing extreme peaks and troughs across
nearly identical outcomes. As the measure of model instability
increases, probabilities from an unstable FOES model additionally
increase in volatility and provably slide into degeneracy. Section
\ref{implications} then emphasizes the implications of such model
instability, showing that such impropriety can be expected to
numerically hinder maximum likelihood estimation and MCMC-based
simulations. As one potential remedy, suggestions are given or
constraining model parameterizations to avoid the most problematic
regions of the parameter space. Proofs of the main results appear in
Appendix \ref{appendix-instab}.

@@

\subsection{Discrete Exponential Family
Models}\label{discrete-exponential-family-models}

@@

\subsection{Restricted Boltzmann
Machines}\label{restricted-boltzmann-machines}

@@

\(N_\mathcal{H}\) of hidden variables \(\boldsymbol H\) in the joint RBM

@@

\subsection{Deep Learning}\label{deep-learning}

@@

\textbf{Deep belief network (DBN).} A DBN resembles a DBM in that there
are multiple layers of latent random variables stacked in a deep
architecture with no conditional dependence between layers. The
difference between the DBM and DBN models is that all but the last
stacked layer in a DBN are Bayesian networks (see Pearl
\protect\hyperlink{ref-pearl985bayesian}{1985}), rather than RBMs. A
Bayesian network is a probabilistic graphical model that defines
conditional dependence to be directed, rather than undirected (as with
the RBM). Thus for visibles \(X_1, \dots, X_N\) with support
\(\mathcal{X}^N, \mid \mathcal{X} \mid < \infty\), a DBN is also a FOES
model with \(q(N)\) the length of parameter vector is dependent on the
dimension of the visibles because
\(P_{\boldsymbol \theta_N}(\boldsymbol x)>0\) for all
\(\boldsymbol x \in\mathcal{X}^N\). Commonly, as in logistic belief nets
(Neal \protect\hyperlink{ref-neal1992connectionist}{1992}), a ``weight''
parameter is placed on each interaction between visibles,
\(X_1, \dots, X_N\), and the first layer of latent variables,
\(H^{(1)}_1, \dots, H^{(1)}_{N_{(H,1)}}\), satisfying the definition of
a FOES model.

\section{Main Results on Model Instability}\label{instability-results}

We now present a formal definition for instability of FOES models as
well as a simple condition for identifying instability in a FOES model
sequence.

\subsection{A Criterion for Instability}\label{criterion}

To define a measure of instability in FOES models, it is useful to
consider the behavior of data models \(P_{\theta_N}\), again supported
on a set \(\mathcal{X}^N\) of outcomes for
\(\boldsymbol X\equiv \boldsymbol X_N =(X_1,\ldots,X_N)\), in connection
to the sample size \(N\). A relevant quantity to this end is a log-ratio
of extremal probabilities (LREP), defined as
\begin{align}
\label{eq:elpr}
 \REP (\boldsymbol \theta_N)  =  \log \left[\frac{\max\limits_{  \boldsymbol x\in \mathcal{X}^N}P_{\boldsymbol \theta_N}( \boldsymbol x)}{\min\limits_{ \boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}( \boldsymbol x)}\right],
\end{align}
based on maximum and minimal model probabilities. In what follows, the
main idea is that instability, and other negative model features, can be
associated with a FOES model formulation for \(N\) random variables
where the LREP \eqref{eq:elpr} is overly large relative to the sample size
\(N\). That is, a sequence of FOES probability models
\(P_{\boldsymbol \theta_N}\) results in specifying the distribution of
observations \(\boldsymbol X=(X_1,\ldots,X_N)\) for each sample size
\(N \geq 1\) and instability will generally occur among these models
whenever the corresponding LREP \eqref{eq:elpr} grows faster than \(N\).
This leads to the following definition.

@@

This definition of S-unstable is a generalization or reinterpretation of
``unstable'' used in Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011}) by allowing
possibly non-exponential family models (e.g., RBM and DBM models in
Sections \ref{restricted-boltzmann-machines}-\ref{deep-learning} as well
as a potentially increasing number \(q(N)\) of parameters through the
parameter sequence \(\boldsymbol \theta_N\in \mathbb{R}^{q(N)}\). While
this definition differs in form and scope from the original, it does
match that in Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011}) for the
special case of exponential models (cf.~Section
\ref{discrete-exponential-family-models} considered there. Section
\ref{illustrations} provides several examples of unstable models as well
as causes for model instability, where the latter may often be traced to
issues in model form (i.e., data functions) and/or parameterization. We
next describe several potentially undesirable features associated with
S-unstable FOES models.

@@

\subsection{Characterizations and Consequences of
Instability}\label{characterizations-and-consequences-of-instability}

As a basic characteristic, S-unstable FOES model sequences have
extremely sensitive probability structures. One aspect is that small
changes in data configuration can lead to very large changes in
probability. Consider, for example, the quantity given by \[
\DN(\boldsymbol \theta_N) \equiv \max \left\{\log \frac{P_{\boldsymbol \theta_N}(\boldsymbol x)}{P_{\boldsymbol \theta_N}(\boldsymbol x^*)} : \boldsymbol x \text{ }\& \text{ } \boldsymbol x^* \in \mathcal{X}^N \text{ differ in exactly one component}\right\},
\] which represents the biggest log-probability ratio for a one
component change in data outcomes in a FOES model with parameter
\(\boldsymbol \theta_N\). We then have the following result prescribing
the behavior of \(\DN(\boldsymbol \theta_N)\) for S-unstable FOES
models.

@@

Theorem \ref{thm:instab-elpr}(i) is a non-asymptotic result, which
connects to the definition of instability in a FOES model through a
log-ratio of extreme model probabilities \eqref{eq:elpr} being too large
relative to the associated sample size \(N\). If so, Theorem
\ref{thm:instab-elpr}(i) guarantees the FOES model must also exhibit
correspondingly large changes in probability for very small differences
among some data configurations, a property that intuitively captures a
notion of instability. Furthermore, and perhaps more seriously under
Theorem \ref{thm:instab-elpr}(ii), S-unstable models can never have
universally bounded changes in probability among single component
variations in data configurations. While not all one component changes
in data may produce massive changes in probability, unstable models must
have some such data outcomes with this property. As a consequence,
unstable probability structures may exhibit extreme sensitivity through
large peaks and troughs over the sample space.

Additionally, S-unstable FOES model sequences are also connected to
degenerate models, where \emph{degeneracy} involves assigning
essentially all probability to modes within the sample space, which
could potentially represent a small subset among the totality of
outcomes. For perspective, note that differing sizes of the scaled
log-ratio \(\REP(\boldsymbol \theta_N)/N\) from \eqref{eq:elpr} induce a
spectrum of levels of instability/stability and Theorem
\ref{thm:instab-elpr} indicates increasing sensitivity of model
probabilities as \eqref{eq:elpr} increases. Furthermore, as the
instability measure grows and the log-ratio
\(\REP(\boldsymbol \theta_N)/N\) diverges, as in the definition
\eqref{eq:Sun} of S-unstable models, then a FOES model sequence will
become degenerate. Theorem \ref{thm:degenFOES} provides a formal
statement of such degeneracy due to S-instability. For a given
\(0 < \epsilon < 1\), define a \(\epsilon\)-modal set of outcomes as
\begin{equation}
\label{eq:mode}
\mathcal{M}_{\epsilon, \boldsymbol \theta_N} \equiv \left\{\boldsymbol x \in \mathcal{X}^N: \log P_{\boldsymbol \theta_N}(\boldsymbol x) > (1-\epsilon)\max\limits_{\boldsymbol y \in \mathcal{X}^N} \log  P_{\boldsymbol \theta_N}(\boldsymbol y) + \epsilon\min\limits_{\boldsymbol y \in \mathcal{X}^N} \log P_{\boldsymbol \theta_N}(\boldsymbol y) \right\}.
\end{equation}
\BeginKnitrBlock{theorem}
\protect\hypertarget{thm:degenFOES}{}{\label{thm:degenFOES}}For any
arbitrarily small \(0 < \epsilon < 1\), an S-unstable FOES model
sequence \(P_{\boldsymbol \theta_N}\), \(N \geq 1\), for
\(\boldsymbol X_N=(X_1, \dots, X_N)\) satisfies \[
P_{\boldsymbol \theta_N}\left( \boldsymbol X_N\in \mathcal{M}_{\epsilon, \boldsymbol \theta_N}\right) \rightarrow 1 \text{ as } N \rightarrow \infty.
\]
\EndKnitrBlock{theorem}

@@

Connected to degeneracy, S-unstable FOES models may also exhibit
additional kinds of extreme and undesirable sensitivity in probabilities
if model parameters \(\boldsymbol \theta_N\) can further be ``dialed''
between positive and negative values. That is, some FOES models
naturally involve parameter spaces covering a positive-negative spectrum
of parameter possibilities, where the signs of parameters provide a
standard device for increasing or decreasing probabilities of outcomes
in the model formulation. In fact, for many models, the switch of a
parameter sign serves to produce reciprocal probabilities, as outlined
in the following model assumption about parameter sign reversal (PSR).

@@

\BeginKnitrBlock{remark}
\iffalse <span class="remark"><em>Remark. \fi{} Model condition PSR and
Corollary \ref{cor:sign} can be extended to cases where parameter
components
\(\boldsymbol \theta_N =(\boldsymbol \theta_{1,N}, \boldsymbol \theta_{N,2})\)
are not all changed in sign (\(-\boldsymbol \theta_N\)) but, more
generally, are instead altered to another parameter configuration
\(\boldsymbol \theta_{N}^A = (\boldsymbol \theta_{N,1}^A, \boldsymbol \theta_{N,2}^A)\)
involving a switch in sign only among some model parameters
\(\boldsymbol \theta_{N,2}^A = -\boldsymbol \theta_{N,2}\) with
remaining parameters \(\boldsymbol \theta_{1,N}^A\) being arbitrarily
chosen. If a sign change occurs among parameters
(\(\pm \boldsymbol \theta_{N,2}\)) which dominate the probability
structure of the model, then an unstable model under
\(\boldsymbol \theta_N\) will then imply that many more unstable models
exist over a broad spectrum of possibilities for variations
\(\boldsymbol \theta_{N}^A\) of \(\boldsymbol \theta_N\).
\EndKnitrBlock{remark}

@@

Model instability can depend intricately on how functions of parameters
and data \(\boldsymbol X_N=(X_1,\ldots,X_n)\) are combined in the
formulation of the model probabilities, though some general causes may
be identified. As one issue, a broad parameter space (or wide
interpretation of this space) may admit some parameters as technically
valid that have an undue and often undesirable impact on the model
structure for a prescribed data size \(N\). In this case, both the size
and dimension of model parameters can be problematic and induce
instability. In combination to this last point, further causes of
instability may also be traced to the magnitude of statistics in the
model. Potentially massive, and thereby unstable, statistics were the
primary focus of instability studies of Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011}) for certain
discrete exponential models having parameters/statistics of fixed
dimension. However, as shown in the following, bounded statistics may
still lead to instability if the parameter dimension is high. We next
provides some examples to illustrate S-instability in FOES models, which
also suggest some potential strategies for preventing unstable models.

\subsection{Equi-probability Models}\label{equi-probability-models}

@@

\subsection{One-parameter Exponential Models}\label{one-param-exp}

@@

\subsection{Fixed-dimensional Linear Exponential
Models}\label{fixed-dim-exp}

@@

\(P_{\boldsymbol \theta_N}\) denote a RBM-based data model \eqref{eq:RBM2}
for \(N\geq 1\) visible variables
\(\boldsymbol X \equiv \boldsymbol X_N\) derived from
\(\tilde{P}_{\boldsymbol \theta_N}\) as the joint RBM distribution
\eqref{eq:RBM1} of \((\boldsymbol X, \boldsymbol H)\) involving some
number \(N_{\mathcal{H}} \equiv N_{N,\mathcal{H}}\geq 0\) of hidden
variables \(\boldsymbol H \equiv \boldsymbol H_N\) and parameters
\(\boldsymbol \theta_N \equiv (\boldsymbol \theta_N^{\mathcal{V}},\boldsymbol \theta_N^{\mathcal{H}}, \boldsymbol \theta_N^{\mathcal{VH}}) \in\mathbb{R}^{N}\times \mathbb{R}^{N_{\mathcal{H}}} \times \mathbb{R}^{N*N_{\mathcal{H}}}\).

@@

\item the instability measure $\REP(\boldsymbol \theta_N)$ for the marginal model $P_{\boldsymbol \theta_N}$ of  $\boldsymbol X$ satisfies

@@

    \left| \REP(\boldsymbol \theta_N)  - \elt\right| \leq    N_{\mathcal{H}}  \log 2

@@

\subsection{Implications for Simulation}\label{mcmc}
